# R Basics and Fundamentals

## Variables {.unnumbered}

Assigning names to values is done in R with the assignment operator: `<-` (as opposed to the conventional `=` which is used with `named function assignments`.)

```{r}
# variable assignment 
x <- 4
y <- x^2 + 2 + x
```

```{r}
# to output what y is (assignment does not print output)
print(y)

# or simply just
y
```

Note: R is a *dynamic language*, so the types and values can easily be changed. (type checks are done during run-time)

```{r}
# x is now a of type character
x <- "statistics"
```

### R data types

-   `numeric (double or int)` - any number with or without a decimal point (and can contain `NA`)

-   `character` - any grouping of characters or string

-   `logical/boolean` - TRUE or FALSE

-   `vectors` list of related data that is all the same type

Use `class()` to determine variable type

```{r}
class(2)
class(x)
```

## Data Vectors

**Vector:** a collection of observations or measurements concerning a single variable

In 2021, the average temperature in Guam for each month was 80.7 81.4 81.5 82.6 82.6 83.7 83.4 81.8 82.5 81.4 82.0 81.0.

```{r}
# Let's place this list of temperatures in a data vector.
# creating a vector - numeric vector
temp <- c(80.7, 81.4, 81.5,82.6, 82.6, 83.7, 83.4, 81.8, 82.5, 81.4, 82.0, 81.0)
# this returns the length of the vector
length(temp)
```

```{r}
sum(temp)
mean(temp)
```

In the case one of your values in your vector is of `NA`, then `sum` or `mean` would also return `NA`. `sum(, na.rm=TRUE)`

**Attributes:** Vectors can have attributes, or names for each case.

```{r}
# assign month to each temperature in the temp vector
months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "July", "Aug", "Sep", "Oct", "Nov", "Dec")

names(temp) <- months
```

## Indexing

You can call a specific value in the `temp` vector by referencing it's respective name.

```{r}
temp["May"]
```

OR, you can index the vector by referencing its numeric index value.

```{r}
# returns the first element in temp
temp[1]
```

```{r}
# if I want the first 4 values, index with :
temp[1:4]
```

## Operators

Comparison:

-   `>` Greater than

-   `<` Less than

-   `<=` Less than or equal

-   `>=` Greater than or equal

-   `==` Equal to:

-   `!=` Not equal to

-   `%in%` in (applicable to vectors)

```{r}
# get all temperatures that are higher than the temperature in Aug
# temp['Aug] refers to temperature in August
# adding temp > temp['Aug] inside temp[] means within the temp vector
temp[temp > temp['Aug']]
```

Logical Operators:

-   `&` AND

-   `|` OR

-   `!` NOT

```{r}
# want temperatures that are greater than temps in August, but also temperatures in the Fall

fall <- c("Sep", "Oct", "Nov")
temp[temp > temp['Aug'] | names(temp) %in% fall]
```
