# R Basics and Fundamentals

## Comments

In `R`, you can comment with `#`. (In markdown, `#` usually denotes the start of a heading.)

```{r}
# comment here
#' 
#' adding ' after #, allows you to enter multi-line comments
#' 

# basic arithmetic
 1+1
```

## Directories

R sessions usually have a *working directory* associated with it. This is the default location where files are imported or saved.

```{r}
# check your working directory 
getwd()
```

```{r eval=FALSE}
# change your working directory 
setwd("/insertfilepathhere")
```

## R Packages

Base installation consists of many in-house functions and commands, but more specialized techniques would require the installation of packages. A few packages that you should have for this course (there will probably be many more):

-   `ggplot2`

-   `data.table`

-   `MASS`

-   `stats`

*Note:* If you are using a mac, some development packages would require you to install `XQuartz` or `XCode`. If issues come up with installing a package or getting a package to work, please let me know so I can help you out!

Find more packages here: <https://cran.r-project.org/>

```{r eval=FALSE}
# installing packages
install.packages("MASS")
```

```{r eval=FALSE}
# calling the package after installation
library("MASS")
```

```{r eval=FALSE}
# updating packages/all packages
update.packages()
```

## Coding

## Variables {.unnumbered}

Assigning names to values is done in R with the assignment operator: `<-` (as opposed to the conventional `=` which is used with `named function assignments`.)

```{r}
# variable assignment 
x <- 4
y <- x^2 + 2 + x
```

```{r}
# to output what y is (assignment does not print output)
print(y)

# or simply just
y
```

Note: R is a *dynamic language*, so the types and values can easily be changed. (type checks are done during run-time)

```{r}
# x is now a of type character
x <- "statistics"
```

### R data types

-   `numeric (double or int)` - any number with or without a decimal point (and can contain `NA`)

-   `character` - any grouping of characters or string

-   `logical/boolean` - TRUE or FALSE

-   `vectors` list of related data that is all the same type

Use `class()` to determine variable type

```{r}
class(2)
class(x)
```

## Data Vectors

**Vector:** a collection of observations or measurements concerning a single variable

In 2021, the average temperature in Guam for each month was 80.7 81.4 81.5 82.6 82.6 83.7 83.4 81.8 82.5 81.4 82.0 81.0.

```{r}
# Let's place this list of temperatures in a data vector.
# creating a vector - numeric vector
temp <- c(80.7, 81.4, 81.5,82.6, 82.6, 83.7, 83.4, 
          81.8, 82.5, 81.4, 82.0, 81.0)
# this returns the length of the vector
length(temp)
```

```{r}
sum(temp)
mean(temp)
```

In the case one of your values in your vector is of `NA`, then `sum` or `mean` would also return `NA`. `sum(, na.rm=TRUE)`

**Attributes:** Vectors can have attributes, or names for each case.

```{r}
# assign month to each temperature in the temp vector
months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
            "July", "Aug", "Sep", "Oct", "Nov", "Dec")

names(temp) <- months
```

## Indexing

You can call a specific value in the `temp` vector by referencing it's respective name.

```{r}
temp["May"]
```

OR, you can index the vector by referencing its numeric index value.

```{r}
# returns the first element in temp
temp[1]
```

```{r}
# if I want the first 4 values, index with :
temp[1:4]
```

### Operators

Comparison:

-   `>` Greater than

-   `<` Less than

-   `<=` Less than or equal

-   `>=` Greater than or equal

-   `==` Equal to:

-   `!=` Not equal to

-   `%in%` in (applicable to vectors)

```{r}
# get all temperatures that are higher than the temperature in Aug
# temp['Aug] refers to temperature in August
# adding temp > temp['Aug] inside temp[] means within the temp vector
temp[temp > temp['Aug']]
```

Logical:

-   `&` AND

-   `|` OR

-   `!` NOT

```{r}
# want temperatures that are greater than temps in August, 
# but also temperatures in the Fall

fall <- c("Sep", "Oct", "Nov")
temp[temp > temp['Aug'] | names(temp) %in% fall]
```

What month gives the highest temperature?

```{r}
temp[temp == max(temp)]
```

```{r}
# if you would just like to extract the month
name <- names(temp)[temp==max(temp)]
```

Let's say we want to find the average temperature in the summer

```{r}
# grab only summer months from temp vector
temp[c("Jun", "July", "Aug")]

# OR 
# define summer months 
summer <- c("Jun", "July", "Aug")
#temp[names(temp) %in% summer]
summer_temps <- temp[summer]
```

The average temperature:

```{r}
avg_summer <- mean(summer_temps)
avg_summer
```

## Vector Operations 

With vectors, you can utilize element-wise basic arithmetic.

For example, if we want to add 2 degrees to every temperature in the `temp` data vector, then we can define that as:

```{r}
temp <- temp + 2
```

Let's print it out to see what it looks like

```{r}
temp
```

Now, let's see what the mean looks like again

```{r}
mean(temp)
```

Now, these degrees are given in Fahrenheit, so let's say we want to convert it into Celcius.

The formula for that is:

$$
C = \frac{5}{9}*(t - 32)
$$

where t is the temperature in Fahrenheit.

```{r}
temp_celcius <- (5/9)*(temp-32)
```

Print out temp celcius (notice it still kept the attributes from temp)

```{r}
temp_celcius
```

```{r}
mean(temp_celcius)
```

## Sequences and Repetition 

R has in-house functions that allow you to create a sequence of values (or repeat).

We will be using `seq` and `rep`, and you can read up for help on the documentation by using ?function in R. This pulls up the documentation and example usage

### Sequences

Without the function, the easiest way to create a sequence with intervals of 1 is using the colon operator

```{r}
1:10
```

For more flexible sequences, let's use `seq`

```{r}
seq(from = 1, to=10, by=3)
```

Note: this will always include the `from` number, but not the `to` number depending on what you're `by` number is.

Instead of specifying a `by` number, you can specify how many numbers you would like in between with `length.out` - this would be equally spaced.

```{r}
seq(from = 1, to = 10, length.out = 20)
```

For decreasing, set `` `by` `` to be a negative value and switch the `from` and `to` values.

```{r}
seq(from = 10, to = 1, by=-3)
```

```{r}
seq(from = 10, to = 1, length.out = 12)
```

### Random Sampling

The `sample()` function takes a sample from the specified elements of `x` with or without replacement.

Create a random sample of 20 from 1 to 100 WITH replacement:

```{r}
sample(1:100, 20, replace = T)
```

Note, running this again won't give you the same output.

```{r}
sample(1:100, 20, replace = T)
```

We avoid this by using `set.seed(x)` where `x` is just any arbitrary number

```{r}
set.seed(415)
sample(1:100, 20, replace = T)
```

```{r}
# run again
set.seed(415)
sample(1:100, 20, replace = T)
```

### Repetition 

In some cases, you may just want to repeat values. We will do this using the `rep` function.

```{r}
# repeat 1 four times
rep(x=1, times = 4)
```

OR

```{r}
# repeat 1 to 5, 10 times
rep(1:5, times = 10)
```

OR

```{r}
# repeat 1, 2, 5, 7, 3 times
rep(x=c(1, 2, 5, 7), times=3)
```

OR

```{r}
# repeat 1, 2, 5, 7 each values twice
rep(x=c(1,2,5,7), times=3, each=2)
```

```{r}
# repeat each element in x by specified vector time
# (1 once, 2 four times, 5, once, and 7 twice)
rep(x=c(1, 2, 5, 7), times=c(1,4,1,2))
```
