# Introduction to Dataframes & ggplot2

#### What is a `data.frame`?

A `data.frame` in R is a data structure used for storing data in a tabular format. It is very similar to a spreadsheet, excel sheet, or SQL table within R. You can also think of it as a collection of multiple data vectors. It's two dimensional, meaning it has rows and columns where each row represents an observation and each column represents a variable, attribute, or feature.

#### What is `tidyverse?`

I will introduce the concept of wrangling, cleaning, and transforming dataframes with the `tidyverse` package. `Tidyverse` is a collection of packages to help with cleaning, transformation, visualization, and analysis. (A few packages in `Tidyverse` include `dplyr`, `ggplot2`, `tidyr`, and `readr`).

#### What is `ggplot2?`

This is *the* R package for data visualization. This suite provides an easy way to visualize your data with: bar charts, scatter plots, time trends, etc. Writing ggplot code comprises of these components: the data being visualized, the aesthetics (how you would like the variables to be mapped), and geometric objects (the shapes that represent the data).

## Case Study: Overwatch 2 Player Statistics

#### Load & Inspect Data

Let's call in our dataset using `read_csv`.

```{r}
library(tidyverse)
# i am defining a string variable with the filepath of the directory that hosts the csv file we are using
dir <- "/Users/rdominguez/Documents/MA500/Lessons/datasets/"

# i use paste0() to concatenate these strings together to provide the full filepath
ow <- read_csv(paste0(dir, "overwatch_stats.csv"))

```

Once you have your data loaded into R, the best first steps to take is to inspect your data. This can look like a lot of things from plotting all your variables to visualize its relationship to other variables in the dataset. But, the main focus is to get familiar with the data and summarize its main characteristics.

```{r}
# using the head() function will bring up the first 6 rows
head(ow)
```

We can also use the `summary()` function on the `ow` data.frame to provide us a summary of each column.

```{r}
summary(ow)
```

`summary()` is most helpful when you want to see the range of values for our numeric variables. When we have categorical or qualitative variables, I like to use the `unique()` function to show me the values these columns provide.

```{r}
# this will output all unique values in the Player column
unique(ow$Player)
```

Our `ow` data.frame includes a `Date` column, but in our `summary()` output, it is categorized as a `character`. Let's re-assign it's class by using the `as.Date` function.

```{r}
# because the values of Date appear as MM/DD/YY, we use
# %m/%d/%y to follow the same format
# %y represents year using 2 digits, where %Y represents 
# year using 4 digits
ow$Date <- as.Date(ow$Date, format="%m/%d/%y")
```

#### Visualize Data

We have a few numeric columns, `elimintations`, `assists`, `deaths`, `damage`, `heals`, and `mitigation`.

Let's plot a histogram for the `damage` column to determine what the distribution of values appear to be.

```{r}
# hist is a base R function
hist(ow$Damage) 
```

We can also create a boxplot to easily determine what our potential outliers look like.

```{r}
# boxplot is also a base R function
boxplot(ow$Damage)
```

### Data Manipulation

Let's say we would like to compare each player's wins. We'll compare wins by plotting a bar chart of wins by Date. It is most effective and straightforward to work with data when it is in `long` format rather than `wide`. (If you take a look at the`ow` dataframe, it's raw state is in `wide`). So, let's make that conversion! To switch from `long` to `wide`, the function is `pivot_wider`, and to go from `wide` to `long`, we use `pivot_longer`.

```{r}
ow <- ow %>% pivot_longer(cols = c("Elimination", "Assists", "Deaths", "Damage", "Heals", "Mitigation"), names_to = "Value_Type", values_to = "Value" )
```

-   `cols` indicates which columns you want to pivot into the long format

-   `names_to` indicates the name of the column you are saving the previous columns

-   `values_to` indicates the name of the column you are saving the values corresponding to the columns in `cols`

Use the `head()` function to see what the first 6 rows look like now!

```{r}
# notice that we now have a column named Value_Type and the values in that column correspond to our previous columns!
head(ow)
```

Next, we want to calculate each player's total number of wins in this dataset, summarized by date. The "Win" column is binary, with 0 indicating a loss and 1 indicating a win. We can simply sum this column for each player to get the total wins.

With `dplyr`, he pipe operator, `%>%`, helps with the readability on the action you are taking with the data.frame. It pipes the value on its left and in the argument that it follows.

```{r}
# we want to group by Player and Date 
# n() is an in house function for tidyverse and simply returns the total within Player and Date
player_wins <- ow %>% group_by(Player, Date) %>% summarise(wins = sum(Win), 
                                                           pct_win = sum(Win)/n()*100)
```

### Plotting with ggplot2

#### Creating a bar chart

```{r}
bar <- ggplot(data = player_wins, aes(x=Date, y=wins))+
  geom_bar(position="dodge", stat="identity", aes(fill = Player))
print(bar)
```

#### Creating a line graph

```{r}
line <- ggplot(data = player_wins, aes(x=Date, y=pct_win))+
  geom_point()+
  geom_line(aes(color = Player))
print(line)
```

Now, let's do the same to determine the number of eliminations each player has based on their role. To achieve this, we'll aggregate the wins by grouping the data by both Player and Role.

```{r}
role <- ow %>% group_by(Player, Role, Value_Type) %>% summarise(val = sum(Value))
head(role)

```

Let's create a barchart for this new aggregation.

```{r}
# we set facets to create multiple barcharts by Value Type
# we also set the scales of the y-axis to be free, meaning
# each barchart will have its own y-axis. we do this
# because the damage and heal values are way larger than 
# the values for deaths, eliminations, assists, etc.
chart <- ggplot(data = role, aes(x=Role, y=val))+
  geom_bar(position="dodge", stat="identity", aes(fill=Player))+facet_wrap(~Value_Type, 
                                                                           scales = "free_y")
print(chart)

```

I want to reorder the x-axis of this chart so that it displays "Tank," "Damage," and then "Support." Additionally, I would like to reorder the facets in the following order: "Eliminations," "Assists," "Deaths," "Damage," "Heals," and "Mitigation." We can achieve this by setting the relevant columns as factors and defining the desired levels for each.

```{r}
# we set the column role to be a factor and define the ordered values with the level argument, we also do the same 
# for the column value type in order to get the facets in our barchart above in the order we would like
role$Role <- factor(role$Role, levels=c("Tank", "Support", "Damage"))
role$Value_Type <- factor(role$Value_Type, levels = c("Elimination", "Assists", "Deaths", "Damage", "Heals", "Mitigation"))
```

Plot the chart again for it to work

```{r}
# plot chart again
chart <- ggplot(data = role, aes(x=Role, y=val))+
  geom_bar(position="dodge", stat="identity", aes(fill = Player))+facet_wrap(~Value_Type, scales = "free_y")+labs(x="Role", y="Number")+theme_minimal()
print(chart)
```

Save the plot with `ggsave`

```{r}
# saving chart 
# if you don't specify a file path, it will save it working directory
ggsave(filename = "/Users/rdominguez/Documents/MA500/chart1.png", 
       plot = chart, 
       dpi = 300)
```

## Merging

Merging data frames is the process of combining two or more data frames based on common columns or row indices. This is a common operation in data analysis, where you often need to bring together information from different sources into a single data set.

In the Overwatch dataset, there's an accompanying dataset that lists the heroes played in each game. We know that each row in the original Overwatch dataset we've been working with corresponds to a single game. Because of the game's mechanics, players are allowed to switch heroes during a match. As a result, in the heroes dataset, you'll see that a player may have played multiple heroes within the same game.

Let's read in this accompanying dataset.

```{r}
hero <- read_csv(paste0(dir, "overwatch_heros.csv"))
hero$Date <- as.Date(hero$Date, format="%m/%d/%y")
```

As you can see, the hero data frame contains more observations than the Overwatch dataset, which has 87 entries.

When you use the `left_join` function to merge the `ow` data frame with the `hero` data frame, it will match all key values from the `hero` data to the `ow` data. However, the first time you run this, you'll encounter a "many-to-many relationship" warning. This means that the join will occur, but it will generate all possible combinations of our key columns—Player, Date, Win, Map, Mode, and Role—with the various corresponding heroes.

To simplify things, we'll only join the rows associated with the heroes that were played the most during each game.

```{r}
# we will first work with filtering the hero data.frame
# this code groups the hero data.frame by our key (unique 
# idenitifying columns) and takes the higher value from the
# Percent Played column

hero_most_played <- hero %>% group_by(Player, Date, Win, Map, Mode, Role) %>% filter(`Percent Played` == max(`Percent Played`))

# left join 
ow_with_hero <- left_join(ow, hero_most_played, by=c("Player", "Date", "Win",
                                                     "Map", "Mode", "Role"))

```

Another warning, a "many-to-many relationship" warning, will appear due to duplicates in the grouped combinations of our key indicators: Player, Date, Win, Map, Mode, and Role. Typically, this issue would be avoided by assigning a unique identifier to each game, but this dataset lacks such an identifier. Essentially, a player can participate in multiple games on the same day, and even play on the same map with the same role and mode. To proceed with our analysis, we'll need to identify these duplicate entries and remove them from the dataset.

```{r}
# we will create a new data.frame called duplicate to count
# the number of instances the grouped key columns appear
duplicate <- ow_with_hero %>% count(Player, Date, Win, Map, Mode, Role, Value_Type) %>% 
  filter(n >1) %>% select(-n) 
```

```{r}
# now we will use an anti-join to remove these duplicate keys 
# from the overwatch dataset 
ow_with_hero <- ow_with_hero %>% anti_join(duplicate, 
                                           by=c("Player", "Date", "Win", 
                                                "Map", "Mode","Role", 
                                                "Value_Type"))

```

The `ow_with_hero` data frame should now have one row for each game, with only one hero per game.

Suppose we want to analyze Player 2's statistics (such as the number of eliminations, assists, heals, damage, deaths, and mitigation) by hero. To do this, we'll need to filter the `ow_with_hero` data frame to include only Player 2's records, and then aggregate or sum the `Value` column to get the total for each hero, regardless of the date.

```{r}
player2 <- ow_with_hero %>% filter(Player == "player2") %>% group_by(Player, Role, Hero, Value_Type) %>% summarise(sum_val = sum(Value))
```

Continue to create a barchart

```{r}
p2 <- ggplot(data = player2, aes(x=Role, y=sum_val))+
   geom_bar(position="dodge", stat="identity", aes(fill = Value_Type))+
   facet_wrap(~Hero, scales="free_y")+labs(title = "Player 2 Statistics by Hero", 
                                                 x = "Role", 
                                                 y = "Number")+theme_minimal()
```

Now, let's output a `.csv` file that contains player statistics (for both player) by hero. For publication or easy reading purposes, it is best to show long datasets and format it to wide.

First, we need to summarise the hero stats again!

```{r}
# this is similar to what we did to get player2 stats above but 
# instead we group by hero, if we filter the output of this
# to just player 2, we should get the same observations as the
# player2 data.frame
hero_stats <- ow_with_hero %>% group_by(Player, Role, Hero, Value_Type) %>% summarise(sum_val = sum(Value))
```

Pivot from long to wide

```{r}
hero_stats <- pivot_wider(hero_stats,
                          id_cols = c("Player", "Role", "Hero"), 
                          names_from = "Value_Type", 
                          values_from = "sum_val")
```

Here I am defining the following parameters:

-   `id_cols`: The key columns that should remain unchanged during pivoting, which are Player, Role, and Hero.

<!-- -->

-   `names_from`: Specifies the column that contains the entries we want to transform into column headers when converting the dataset to a wide format.

-   `values_from`: Specifies the column from which we want to extract the cell values for the newly created columns.

To save this to a `.csv` file, we will use the `write_csv` function.

```{r}
# first argument is the data.frame you are saving
# second argument is the filepath we would like to save it to
write_csv(hero_stats, paste0(dir, "ow_hero_stats.csv"))
```
