# Data Visualizations

Beyond the bar charts and line graphs discussed in the previous section, ggplot2 offers a variety of other visualizations that you can create. We'll be using the `mtcars` dataset, which is readily available in your R environment, to explore these options.

```{r}
data(mtcars)
head(mtcars)
```

## Types of Plots

### Scatter Plot

-   Shows relationship between two continuous variables

```{r}
scatter <- ggplot(data = mtcars, 
                  aes(x = wt, y = mpg))+
  geom_point()+
  labs(x = "Weight",
       y = "Miles per Gallon")
print(scatter)
```

### Line Plot

-   Displays trends over time or ordered data

```{r}
line <- ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_line() +
  labs(x = "Horsepower",
       y = "Miles Per Gallon")
print(line)

# if you want to plot the trend of the different
# values in a categorical variable, you can 
# set it as a color 
line <- ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_line(aes(color = factor(gear))) +
  labs(x = "Horsepower",
       y = "Miles Per Gallon", 
       color = "Number of Gear")
print(line)
```

### Bar Plot

-   Displays counts or summary statistics of categorical data

```{r}
bar <- ggplot(mtcars, aes(x = factor(cyl))) +
  geom_bar(fill = "steelblue") +
  labs(x = "Number of Cylinders",
       y = "Count")
print(bar)
```

### Histogram

-   Displays the distribution of a single continuous variable

```{r}
hist <- ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 2, fill = "steelblue") +
  labs(title = "Histogram of MPG",
       x = "Miles Per Gallon",
       y = "Frequency")
print(hist)
```

### Boxplot

-   Displays the distribution of one or multiple continuous variables

```{r}
# you can fill/color the boxes to correspond to a # specific categorical variable as well 
box <- ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot(aes(fill = factor(cyl))) +
  labs(title = "Box Plot of MPG by Cylinder",
       x = "Number of Cylinders",
       y = "Miles Per Gallon", 
       fill = "Number of Cylinders")
print(box)
```

### Density Plot

-    Displays the distribution as a smoothed curve of a continuous variable

```{r}
density <- ggplot(mtcars, aes(x = mpg)) +
  geom_density(fill = "blue", alpha = 0.5) +
  labs(x = "Miles Per Gallon",
       y = "Density")
print(density)

```

### Violin Plot

-    To show the distribution of a continuous variable by categories, combining aspects of a box plot and a density plot.

```{r}
violin <- ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_violin(aes(fill = factor(cyl))) +
  labs(x = "Number of Cylinders",
       y = "Miles Per Gallon", 
       fill = "Number of Cylinders")
print(violin)
```

### Faceted Plots

-   Creates multiple plots based on the values of a categorical variable

```{r}
facet <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(aes(color = factor(cyl))) +
  facet_wrap(~ cyl) +
  labs(x = "Weight",
       y = "Miles Per Gallon (mpg)", 
       color = "Number of Cylinders")
print(facet)
```

## Themes

`ggplot2`offers a variety of built-in themes that you can apply to change the appearance of your plots. Here are a few options you can use:

-   `theme_grey()`

-   `theme_bw()`

-   `theme_linedraw()`

-   `theme_light()`

-   `theme_dark()`

-   `theme_minimal()`

-   `theme_classic()`

-   `theme_void()`

Here is how you can implement:

```{r}
# re-using the box plot, we can append a new
# ggplot element to our variable box

box <- box + theme_minimal()
print(box)
```
